---
title: "A Descriptive Report of the NHANES Dataset"
author: "Mushabe Brian"
date: "2026-10-04"
output: pdf_document
bibliography: references.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

##Loading packages 
```{r}
library(tidyverse)
library(knitr)
library(kableExtra)
```

# Introduction

The National Health and Nutrition Examination Survey (NHANES) is a program of
studies designed to assess the health and nutritional status of people in the
United States. This report describes the cleaned NHANES sample by survey wave,
including sample size, sex, ethnicity, and average systolic blood pressure.

# Data preparation

The code below assumes the data have columns named `wave`, `sex`, `ethnicity`,
and `average_systolic_bp`. Replace these names and category labels as needed.
The percentages for sex and ethnicity use the total sample size within each
wave as the denominator. Missing values are excluded from the relevant category
counts and blood-pressure summaries.

## import data 
```{r}
pacman::p_load(here, rio)
file.copy(
  from = "C:/Users/userx/Downloads/cleaned_NHANES (1).csv",
  to = here("data", "cleaned_nhanes.csv"),
  overwrite = TRUE
)
cleaned_nhanes <- import(here("data", "cleaned_nhanes.csv"))
```

# Table 1: Overall Sample Size and Male Demographics by Wave
```{r}
table1_data <- cleaned_nhanes %>%
  group_by(wave_id) %>%
  summarise(
    Total_Sample_Size = n(),
    Male_Count = sum(tolower(gender) == "male", na.rm = TRUE),
    Male_Percent = round((Male_Count / Total_Sample_Size) * 100, 1)
  )
names(cleaned_nhanes)
```


## ethinicity of the participants 
```{r}
table2_data <- cleaned_nhanes %>%
  group_by(wave_id, ethnicity_1) %>%
  summarise(Count = n(), .groups = "drop") %>%
  group_by(wave_id) %>%
  mutate(Percent = round((Count / sum(Count)) * 100, 1)) %>%
  pivot_wider(names_from = wave_id, values_from = c(Count, Percent))
```

## systolic blood pressure measurements 
```{r}
# 1. Compute the average systolic BP for each patient across the available readings
cleaned_nhanes <- cleaned_nhanes %>%
  mutate(avg_systolic = rowMeans(select(., systolic_bp_1, systolic_bp_2, systolic_bp_3, systolic_bp_4), na.rm = TRUE))

# 2. Summarize summary statistics by wave_id
table3_data <- cleaned_nhanes %>%
  group_by(wave_id) %>%
  summarise(
    Min_BP = min(avg_systolic, na.rm = TRUE),
    Max_BP = max(avg_systolic, na.rm = TRUE),
    Median_BP = median(avg_systolic, na.rm = TRUE),
    Mean_BP = round(mean(avg_systolic, na.rm = TRUE), 1),
    SD_BP = round(sd(avg_systolic, na.rm = TRUE), 1)
  )
view(table3_data)
```

## advanced formatting options to emphasize information of your choice.
```{r}
table(cleaned_nhanes$ethnicity_1)
table(cleaned_nhanes$avg_systolic, cleaned_nhanes$wave_id)
```

```{r}
# Create contingency table of wave and gender
gender_wave_table <- xtabs(~ wave_id + gender, data = cleaned_nhanes)

# Add row and column totals
gender_wave_margins <- addmargins(gender_wave_table)

# Convert to a data frame for kable formatting
table1_df <- as.data.frame.matrix(gender_wave_margins)
gender_wave_table
```
```{r}
# Create contingency table for wave and ethnicity_1
eth_wave_table <- xtabs(~ ethnicity_1 + wave_id, data = cleaned_nhanes)

# Calculate column proportions, multiply by 100, and round to 2 decimals
eth_pct_table <- round(prop.table(eth_wave_table, margin = 2) * 100, 2)

# Combine counts and percentages if desired, or format directly
table2_df <- as.data.frame.matrix(eth_wave_table)
eth_wave_table
```
##Sample Size and Gender Distribution

```{r}
kable(
  gender_wave_margins, 
  format = "latex",
  booktabs = TRUE,
  col.names = c("Wave", "Female", "Male", "Total"),
  caption = "Overall Sample Size and Gender Distribution by Wave"
) %>%
  kable_styling(
    latex_options = c("striped", "hold_position"),
    full_width = FALSE
  ) %>%
  row_spec(0, bold = TRUE) %>%                    
  column_spec(1, bold = TRUE) %>%                   
  column_spec(ncol(gender_wave_margins), bold = TRUE)
```

## Ethnicity Breakdown Across Waves
Understanding the ethnic diversity of the sample is critical for population health assessments. Below, cross-tabulation and row/column proportions (prop.table()) highlight the distribution of participants across different ethnic categories per wave.
```{r}
kable(
  eth_wave_table, 
  format = "latex",
  booktabs = TRUE,
  col.names = c("Ethnicity", "Wave 1 (n)", "Wave 2 (n)"), # Adjust column names to match your wave columns
  caption = "Ethnicity Breakdown Across Survey Waves"
) %>%
  kable_styling(
    latex_options = c("striped", "hold_position", "scale_down"),
    full_width = FALSE
  ) %>%
  row_spec(0, bold = TRUE) %>%
  column_spec(1, bold = TRUE)
```


## Systolic Blood Pressure Summary Statistics
Clinical measurements such as blood pressure provide vital benchmarks for cardiovascular health. Participants underwent multiple blood pressure measurements (systolic_bp_1 through systolic_bp_4). We compute each participant's average systolic reading, then aggregate summary statistics (dplyr grouped by wave_id).
```{r}
kable(
  table3_data, 
  format = "latex",
  booktabs = TRUE,
  col.names = c("Wave", "Minimum", "Maximum", "Median", "Mean", "Std. Dev."),
  caption = "Summary Statistics of Average Systolic Blood Pressure by Wave"
) %>%
  kable_styling(
    latex_options = c("striped", "hold_position"),
    full_width = FALSE
  ) %>%
  row_spec(0, bold = TRUE) %>%
  column_spec(1, bold = TRUE) %>%
  column_spec(5, bold = TRUE) # Emphasizing the Mean column
```

















