---
title: "Exercise 2-4"
author: "Mushabe Brian"
date: "2026-10-04"
output: pdf_document
---

```{r setup, include=FALSE}
okabe_ito_palette <- c(
  "#E69F00", "#56B4E9", "#009E73", "#F0E442",
  "#0072B2", "#D55E00", "#CC79A7", "#000000"
)
knitr::opts_chunk$set(echo = TRUE)
```

# ==========================================
# Exercise 2-Visualizing key characteristics
# ==========================================
```{r}
##Loading the required packages 
library(tidyverse)
library(here)
library(rio)
```

```{r}
##create the required folders in the working directory
dir.create(here("data"), showWarnings = FALSE)
dir.create(here("scripts"), showWarnings = FALSE)
dir.create(here("figures"), showWarnings = FALSE)
dir.create(here("reports"), showWarnings = FALSE)
```


```{r}
# Install/load packages as needed
pacman::p_load(here, rio)

# Ensure the repository's data folder exists
dir.create(here("data"), showWarnings = FALSE, recursive = TRUE)

# Copy the source files into the repository's data folder
file.copy(
  from = "C:/Users/userx/Downloads/cleaned_NHANES (1).csv",
  to = here("data", "cleaned_nhanes.csv"),
  overwrite = TRUE
)

file.copy(
  from = "C:/Users/userx/Downloads/diet (1).csv",
  to = here("data", "diet (1).csv"),
  overwrite = TRUE
)

# Import the files after copying them
nhanes_data <- rio::import(here("data", "cleaned_nhanes.csv"))
diet_data  <- rio::import(here("data", "diet (1).csv"))

```

```{r}
##Using ggplot() to create appropriate visualisations 
view(nhanes_data$age)
visual_age <- p_age <- ggplot(nhanes_data, aes(x = age)) +
  geom_histogram(binwidth = 5, fill = "#009E73", color = "white", alpha = 0.8) +
  labs(
    title = "Distribution of Age in the NHANES Dataset",
    subtitle = "Each bin represents a 5-year interval",
    x = "Age (Years)",
    y = "Count"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    panel.grid.minor = element_blank()
  )

visual_age
##checking for misingness 
sum(is.na(nhanes_data$age))
sum(nhanes_data$age)
```
##Distribution of age 
For the variable age, the distribution shows a right-skewed pattern. The age of the patients range from 0 to 80 years with a median age of 28ears and average age of 32.5 years. There are no missing vales in the variable age.


```{r}
visual_gender <- ggplot(nhanes_data, aes(x = gender, fill = gender)) +
  geom_bar(alpha = 0.8) +
  scale_fill_manual(values = c("#0072B2", "#D55E00")) +
  labs(
    title = "Distribution of Gender in the NHANES Sample",
    x = "Gender",
    y = "Count",
    fill = "Gender"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    legend.position = "none",
    panel.grid.minor = element_blank()
  )

visual_gender

```
##Distribution of Gender
Among the patients, (14,948)50.8% are female and 14,452(49.2%) and there are no missing values in the variable gender

```{r}
##Distribution of Ethinicity1
visual_ethinicity_1 <- ggplot(nhanes_data, aes(x = ethnicity_1, fill = ethnicity_1)) +
  geom_bar(alpha = 0.8) +
  scale_fill_manual(values = okabe_ito_palette) +
  labs(
    title = "Distribution of  Ethnicity1",
    x = "Ethnicity 1",
    y = "Count",
    fill = "Ethnicity1"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    axis.text.x = element_text(angle = 30, hjust = 1),
    legend.position = "none"
  )

visual_ethinicity_1
##checking for missingness
sum(is.na(nhanes_data$ethnicity_1))
table(nhanes_data$ethnicity_1)
```

##Distribution of first category of ethnicity1
The total number of participants is 29,400, of these the majority are white 9,890 (33.6%) followed by the black participants 6,511 (22.2%). The lowest category of participants is hispanic 3,088 (10.5%). This category of ethinicity has no missing values.


```{r}
visual_ethinicity_2 <- ggplot(nhanes_data, aes(x = ethnicity_2, fill = ethnicity_2)) +
  geom_bar(alpha = 0.8) +
  scale_fill_manual(values = okabe_ito_palette) +
  labs(
    title = "Distribution of  Ethnicity2",
    x = "Ethnicity 2",
    y = "Count",
    fill = "Ethnicity2"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    axis.text.x = element_text(angle = 30, hjust = 1),
    legend.position = "none"
  )

visual_ethinicity_2
##checking for missingness
ethinicity2_table <- table(nhanes_data$ethnicity_2, useNA = "ifany")

data.frame(
  ethin = names(ethinicity2_table),
  Frequency = as.vector(ethinicity2_table),
  Percentage = round(as.vector(prop.table(ethinicity2_table) * 100), 1)
)
```
```{r}
summary(nhanes_data$age)
table(nhanes_data$gender)
table(nhanes_data$ethnicity_1)
table(nhanes_data$ethnicity_2)
```
```{r}
##Using "ggsave" to save all plots in the figure subfolder in the working directory
##saving age
ggsave(
  filename = "age_distribution_plot.png",
  plot = visual_age,
  path = here("figures"),
  width = 8, 
  height = 6
)
##saving gender
ggsave(
  filename = "gender_distribution_plot.png",
  plot = visual_gender,
  path = here("figures"),
  width = 8, 
  height = 6
)
##saving ethinicity1
ggsave(
  filename = "Ethinicity1_distribution_plot.png",
  plot = visual_ethinicity_1,
  path = here("figures"),
  width = 8, 
  height = 6
)
##saving ethinicity2
ggsave(
  filename = "Ethinicity2_distribution_plot.png",
  plot = visual_ethinicity_2,
  path = here("figures"),
  width = 8, 
  height = 6
)
```

```{r}
##choosing from the two categories of ethinicity
##structure of the two categories 
str(nhanes_data$ethnicity_1)
str(nhanes_data$ethnicity_2)
##tabulating the categoties 
table(nhanes_data$ethnicity_1)
table(nhanes_data$ethnicity_2)
##checking for missingness
sum(is.na(nhanes_data$ethnicity_1))
sum(is.na(nhanes_data$ethnicity_2))
```
##Decision
Ater analysisng the two categories, i decided to keep Ethinicity2 category and drop category 1. Whereas both categories have an equal number of observations (29,400), ethinicity1 has 5 categories while ethinicity has 6 categories. Ethinicity2 adds on the "asian" category which makes it more representaive than category 1. 

```{r}
##keeping Ethinicity
nhanes_data <- nhanes_data %>%
  select(-ethnicity_1)
names(nhanes_data)
```
# ==========================================
# Exercise 3-Improving ggplot figure
# ==========================================
##points that need improvement in your colleague's figure and how i will improve them.
In my colleague's plot, there's no title, i will create a plot and give a title "Participant's Weight Change from Baseline weight". This helps the audience the the objective of the plot at first glance.

The objective of the study is to show how each participants weight had changed over time from the baseline weight. My colleague's plot doesn't show baseline weight. I will create a baseline weight (at 0) and show how each participant deviates from it. Those above it have had a weight gain and those below it have had a weight loss. 

```{r}
## To determine how each participant's weight has changed over time in comparison to the baseline weight 
diet_data <- import(here("data", "diet (1).csv"))

change_in_weight <- diet_data %>%
  group_by(Participant) %>%
  arrange(Week) %>%
  mutate(
    baseline_weight = Weight[Week == 0][1],
    change_from_baseline = Weight - baseline_weight
  ) %>%
  ungroup()

head(change_in_weight)
change_in_weight
```

From the results, 0 means there was no change in weight after following the diet, negative means theres was a loss in weight and a positive means there was a gain in weight. 

```{r}
##Plotting the right spaghetti plot to show each participants weight has changed 
library(RColorBrewer) # Required for ColorBrewer palettes

plot_data <- change_in_weight %>%
  group_by(Participant) %>%
  arrange(Week, .by_group = TRUE) %>%
  mutate(
    baseline_weight = first(Weight[Week == 0]),
    change_from_baseline = Weight - baseline_weight
  ) %>%
  ungroup()

# 1. Generate a color palette using RColorBrewer
num_participants <- n_distinct(plot_data$Participant)
my_colors <- colorRampPalette(brewer.pal(8, "Set2"))(num_participants)

# 2. Build the plot with individual distinct colors
ggplot(
  plot_data,
  aes(
    x = Week,
    y = change_from_baseline,
    group = Participant,
    color = factor(Participant) # Map a unique color per participant
  )
) +
  # Reference line at zero (Baseline)
  geom_hline(
    yintercept = 0,
    linetype = "dashed",
    color = "black",
    linewidth = 0.7
  ) +
  # Individual trajectories
  geom_line(
    linewidth = 1,
    alpha = 0.8
  ) +
  geom_point(
    size = 1.6,
    alpha = 0.8
  ) +
  # Apply the RColorBrewer custom palette
  scale_color_manual(values = my_colors) +
  # X-axis formatting
  scale_x_continuous(
    breaks = seq(0, 16, by = 2)
  ) +
  # Professional labels and titles
  labs(
    title = "Participant's Weight Change from Baseline weight",
    x = "Week on diet",
    y = "Change from baseline weight (kg)",
    color = "Participant"
  ) +
  # Theme styling
  theme_minimal(base_size = 13) +
  theme(
    plot.title = element_text(face = "bold"),
    legend.position = "right" # Set to "none" if the legend is too large for your layout
  )
```

# ==========================================
# Exercise 4-Exploring relationships through visualizations
# ==========================================
##The objective of this part is to explore relationships between age, gender, and systolic blood pressure
##(SBP) using effective visual storytelling
```{r}
names(nhanes_data)
view(nhanes_data)
##Averaging the systolic blood pressure variables 
names(nhanes_data)
nhanes_data <- nhanes_data %>%
  mutate(
    avg_sbp = rowMeans(
      select(., systolic_bp_1, systolic_bp_2, systolic_bp_3, systolic_bp_4), 
      na.rm = TRUE
    ),
    avg_sbp = ifelse(is.nan(avg_sbp), NA, avg_sbp)
  )
##Create a new categorical variable that classifies individuals based on their average SBP
nhanes_data <- nhanes_data %>%
  mutate(
    sbp_category = case_when(
      avg_sbp < 120 ~ "Normal",
      avg_sbp >= 120 & avg_sbp < 130 ~ "Elevated",
      avg_sbp >= 130 & avg_sbp < 140 ~ "Stage 1 hypertension",
      avg_sbp >= 140 ~ "Stage 2 hypertension",
      TRUE ~ NA_character_ 
    ),
    sbp_category = factor(
      sbp_category, 
      levels = c("Normal", "Elevated", "Stage 1 hypertension", "Stage 2 hypertension")
    )
  )
names(nhanes_data)
view(nhanes_data)
##Use the filter() function to create a new datasets that includes only individuals who have
#non-missing values for age, gender, and average SBP.
nhanes_clean <- nhanes_data %>%
  filter(!is.na(age_cat) & !is.na(gender) & !is.na(avg_sbp))
view(nhanes_clean)
names(nhanes_clean)
##the new data set (nhanes_clean) has 23 variables and 21,604 observations
##using facet_wrap() to investigate hypertension varies across age and gender 
ggplot(nhanes_clean, aes(x = age_cat, fill = sbp_category)) +
  geom_bar(position = "fill", color = "black", linewidth = 0.3) +
  ##using the facet_wrap to display the relationship
  facet_wrap(~ gender) +
  scale_fill_manual(
    values = c(
      "Normal"              = "#009E73", # Green
      "Elevated"            = "#F0E442", # Yellow
      "Stage 1 hypertension" = "#E69F00", # Orange
      "Stage 2 hypertension" = "#D55E00"  # Red-Orange
    )
  ) +
  labs(
    title = "Distribution of Blood Pressure Categories by Age and Gender",
    x = "Age Category",
    y = "Proportion",
    fill = "SBP Category"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    plot.subtitle = element_text(color = "gray30", size = 11),
    axis.text.x = element_text(angle = 25, hjust = 1), 
    legend.position = "right",
    panel.grid.minor = element_blank(),
    strip.text = element_text(face = "bold", size = 11) 
  )

```

```{r}
##use of 4 plots to describe the final sample size
old_dataset <- nrow(nhanes_data)
new_dataset <- nrow(nhanes_clean)
n_removed <- old_dataset - new_dataset
percent_removed <- 100 * n_removed / old_dataset

sample_sizes <- data.frame(
  Dataset = c("old sample", "new sample"),
  Observations = c(old_dataset, new_dataset)
)

sample_sizes
```

```{r}
##Plot showing hypertension categories 
##defining 
sbp_colors <- c(
  "Normal" = "#2CA25F",
  "Elevated" = "#FDD835",
  "Stage 1 hypertension" = "#FB8C00",
  "Stage 2 hypertension" = "#D7301F"
)

plot_sbp <- ggplot(
  nhanes_clean,
  aes(x = sbp_category, fill = sbp_category)
) +
  geom_bar() +
  scale_fill_manual(values = sbp_colors, drop = FALSE) +
  labs(
    title = "Hypertension categories in the final sample",
    x = "Mean systolic blood pressure category",
    y = "Number of participants"
  ) +
  theme_minimal() +
  theme(
    legend.position = "none",
    axis.text.x = element_text(angle = 25, hjust = 1)
  )

plot_sbp


```

```{r}
##Plot of gender distribution 
p_gender <- ggplot(nhanes_clean, aes(x = gender, fill = gender)) +
  geom_bar() +
  labs(
    title = "Gender distribution in the final sample",
    x = "Gender",
    y = "Number of participants"
  ) +
  scale_fill_brewer(palette = "Set2") +
  theme_classic() +
  theme(legend.position = "none")

p_gender
```

```{r}
##reuse one saved plot from Exercise 2 and compare it to the same variable's distribution in your final
##sample
##plot from exercise 2 was saved as "age_distribution_plot.png"
library(dplyr)
library(ggplot2)

# Recreate the data used for the original saved plot
age_comparison <- bind_rows(
  nhanes_data %>%
    filter(!is.na(age)) %>%
    transmute(
      age,
      sample = "Original sample"
    ),

  nhanes_clean %>%
    filter(!is.na(age)) %>%
    transmute(
      age,
      sample = "Final sample"
    )
)

# Faceted comparison using the same histogram style
age_comparison <- bind_rows(
  nhanes_data %>%
    filter(!is.na(age)) %>%
    select(age) %>%
    mutate(sample = "Original sample"),

  nhanes_clean %>%
    filter(!is.na(age)) %>%
    select(age) %>%
    mutate(sample = "Final sample")
)

names(age_comparison)
##the faceted plot 
age_comparison_plot2 <- ggplot(
  age_comparison,
  aes(x = age)
) +
  geom_histogram(
    binwidth = 5,
    fill = "steelblue",
    color = "white"
  ) +
  facet_wrap(~ sample, ncol = 1) +
  labs(
    title = "Age Distribution original and final distribution",
    x = "Age",
    y = "Number of Participants"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold"),
    strip.text = element_text(face = "bold")
  )

age_comparison_plot2


```

##Comparison of the plots before and after. 
The age distribution generally remained similar before and after. However, the original histogram show more concentration in the early bins and the final shows more flatter bins. This means that those removed because of missing gender, mean SBP may not have been evenly distributed. In conclusion, filtering may have reduced the sample size but it doesnt change the distribution of age. 

## A short report
To analyze systolic blood pressure (SBP), i calculated an average for each participant using up to four available readings. Participants missing all four readings were excluded, along with anyone missing age or gender data. 

The final sample contains fewer participants than the original data set, and its age distribution is not identical to the original distribution. Although the overall age range remains similar, the final sample is flatter at the younger end, where the original data set shows more pronounced bars. This suggests that missing data were not distributed entirely at random across age groups and that filtering may have changed the representation of younger participants.

Younger participants are more commonly classified as having normal or elevated blood pressure, whereas older participants show a greater proportion of Stage 1 and Stage 2 hypertension. These age-related patterns are visible in both genders, although the proportions may differ somewhat between males and females within particular age groups.

Overall, the visualizations suggest an association between increasing age and higher SBP in this cohort. 





